---
title: "limma_analysis"
author: "Hsiao-Yu Peng"
date: "2025-02-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#BiocManager::install("limma")
library(limma)
```

```{r}
# load RMA data
expr_data <- read.table("GSE7621_RMA_normalized.txt", header=TRUE, row.names=1, sep="\t")

head(expr_data)
```



```{r}
# Read metadata to ensure all row.names are correct
metadata <- read.table("GSE7621_metadata.txt", header=TRUE, sep="\t", row.names=1)

# Ensure metadata sample_id is the same as expr_data
metadata <- metadata[colnames(expr_data), ,drop=FALSE]

# label as "factor"
metadata$label <- as.factor(metadata$label)

# Design matrix
design <- model.matrix(~ 0 + metadata$label) 
colnames(design) <- levels(metadata$label)

print(design)
```

```{r}
# Use limma to analyze DEA
fit <- lmFit(expr_data, design)
contrast_matrix <- makeContrasts(PD - Normal, levels = design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

# Get the results
result <- topTable(fit2, adjust="fdr", number = Inf)

# Save the result
write.csv(result, file="limma_DEGs_results.csv")

# Show result
head(result)

```

1. LogFC（Log Fold Change): 
- logFC > 1 (upregulated): Higher expression in PD patients
- logFC < 1 (downregulated): Lower expression in PD patient

2. P.value: original statistical p-value

3. adj.P.Val (adjusted p-value by FDR): Select significant expressed genes, use `adj.P.Val < 0.05`

4. B: B value, the higher, the more confident.

```{r}
# Select significant expressed genes （FDR < 0.05, LogFC > |1|）
DEGs <- result[result$adj.P.Val < 0.05 & abs(result$logFC) > 1, ]
write.csv(DEGs, file = "significant_DEGs.csv")

head(DEGs)

```

## Visualization
### Volcano Plot

```{r}
library(ggplot2)

result$threshold <- as.factor(result$adj.P.Val < 0.05  & abs(result$logFC) > 1)
ggplot(result, aes(x=logFC, y=-log10(P.Value), color=threshold)) +
  geom_point(alpha=0.5) +
  theme_minimal() +
  labs(title = "Volcano Plot", x="log 2 Fold Change", y="-log10 P Value")


```


## Enrichment Analysis

The names in the "Gene" column are in formats like `204457_s_at` or `227321_at`, which indicates that the gene identifiers are **not Gene Symbols** but **Affymetrix Probe IDs**.


**1. Why can't `bitr()` convert these IDs?**

Because `bitr()` requires **standard Gene Symbols** (e.g., `SNCA`, `MAPT`, `LRRK2`), but the format provided **Affymetrix Probe IDs** (e.g., `204457_s_at`).  
These Probe IDs need to be converted into Gene Symbols first before they can be mapped to Entrez IDs.

**2. Convert Probe IDs to Gene Symbols**

In R, we can use `AnnotationDbi` along with the `hgu133plus2.db` package to convert **Affymetrix Probe IDs → Gene Symbols**.  

```{r}
#BiocManager::install(c("clusterProfiler", "org.Hs.eg.db", "ReactomePA", "DOSE"))
#BiocManager::install("hgu133plus2.db")

library(hgu133plus2.db)
library(clusterProfiler)
library(org.Hs.eg.db) # human gene annotation
library(ReactomePA) # Reactome Pathway analysis
library(DOSE) # Disease Enrichment Analysis
```

### Covert Gene Format

Many pathway analysis needs **Entrez ID**, not Gene Symbol.

```{r}
DEGs <- read.csv("significant_DEGs.csv", row.names = 1)

# Obtain Probe_ID list -> Gene column
DEGs$Gene <- rownames(DEGs)
rownames(DEGs) <- NULL # clear out row names to avoid affecting the following analysis

# Convert Probe ID to Gene Symbol
gene_symbols <- mapIds(hgu133plus2.db, keys=DEGs$Gene, column="SYMBOL", keytype="PROBEID", multiVals="first")

# Add Gene Symbol to DEGs
DEGs$GeneSymbol <- gene_symbols

# Remove no those rows without corresponding Gene Symbol (avoid errors later)
DEGs_clean <- na.omit(DEGs)

#write.csv(DEGs_clean, file="significant_DEGs_clean_results.csv")
```


### Covert Gene Sumbol into Entrez ID

```{r}
gene_list <- bitr(DEGs_clean$GeneSymbol, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

head(gene_list)

```


# 4.1 GO Enrichment Analysis

```{r}
go_results <- enrichGO(gene = gene_list$ENTREZID, OrgDb = org.Hs.eg.db, keyType = "ENTREZID",
                       ont = "BP", pAdjustMethod = "fdr", readable = TRUE)

# Create GO enrichment results
barplot(go_results, showCategory=10)
dotplot(go_results, showCategory=10)

```


4.2 KEGG Pathway Enrichment Analysis

```{r}
kegg_results <- enrichKEGG(gene = gene_list$ENTREZID, organism = 'hsa')

# Create KEGG Pathway
barplot(kegg_results, showCategory=10)
dotplot(kegg_results, showCategory=10)
```


# 4.3 Reactome Pathway Analysis

Reactome is a more detail Pathway Database than KEGG, good for exploring detailed molecular mechanism.

```{r}
reactome_results <- enrichPathway(gene = gene_list$ENTREZID, organism = "human")

# Create Reactome Pathway
barplot(reactome_results, showCategory=10)
dotplot(reactome_results, showCategory=10)
```

From GO, KEGG, and Reactome enrichment analyses, we identified several pathways significantly associated with PD, including neurotransmitter-related pathways (e.g., dopamine synapse) and synaptic vesicle processes. These findings are consistent with the known pathophysiology of Parkinson's disease.


